---
title: "Geospatial Data Analysis_P2"
author: "Fan Jing"
format: html
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
editor: visual
---

# 1. Installing and Launching R Packages

```{r}
pacman::p_load(SmartEDA, tidyverse, ggExtra, ggiraph)
```

# 2 Data Wrangling

## 2.1 Importing the data set

In the code chunk below, read_csv() of readr is used to import WDI Country Data.csv into R and parsed it into tibble R data.frame format.

```{r}
water_indicators <- read.csv("WDICountryData.csv")
```

## 2.2 Data summary

Function ExpData() comes from the SmartEDA package in R and The type = 2 argument specifies the type of analysis to perform.

```{r}
water_indicators %>%
  ExpData(type = 2)
```

## 2.3 Data Cleaning

```{r}
water_indicators[6:27] <- lapply(water_indicators[6:27], function(x) as.numeric(as.character(x)))
```

Changing column name

```{r}
water_long <- water_indicators %>%
  pivot_longer(
    cols = starts_with("X"),             
    names_to = "Year",
    values_to = "Value"
  )
```

```{r}
water_long <- water_long %>%
  mutate(Year = gsub("X(\\d{4})\\.\\..*", "\\1", Year),
         Year = as.numeric(Year),
         Value = as.numeric(Value))  
```

First, we will select latest records in 2021 for nine indicator.

```{r}
water_long_selected <- water_long %>%
  filter(Year == 2021)
```

```{r}
water_long_selected[water_long_selected == ".."] <- NA
water_long_selected[water_long_selected == "#N/A"] <- NA
water_long_selected[water_long_selected == ""] <- NA

```

Next, reshape the data for 2021 and change data type

```{r}
water_long_selected_reshape <- water_long_selected %>%
  select(Country = Country.Name, Region, Variable = Series.Name, Value) %>%
  pivot_wider(names_from = Variable, values_from = Value) %>%
  drop_na("Region")
```

```{r}
water_long_selected_reshape$`Population, total` <- as.numeric(gsub(",", "", water_long_selected_reshape$`Population, total`))
```

# 3. Plot Marginal Density Function

## 3.1 Simple scatterplot between Level of water stress (%) and Water Productivity (US\$ GDP per m³) by region

### 3.1.1 African Region

```{r}
water_long_selected_reshape_9 <- water_long_selected_reshape %>%
  select('Country', 'Region',`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  filter(Region == 'African Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_9 <- water_long_selected_reshape_9 %>%
  drop_na(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
          `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  mutate(
    `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources` = 
      round(as.numeric(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`), 0),
    
    `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)` = 
      round(as.numeric(`Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`), 0)
  )

```

```{r}
p1 <- ggplot(data = water_long_selected_reshape_9, 
             aes(x = `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
                 y = `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(0, 140), ylim = c(0, 140)) +
  labs(x = "Level of water stress (%)", 
       y = "Water Productivity (US$ GDP per m³)",
       color = "Country Name")
p1
```

```{r}
girafe(ggobj = p1)
```

### 3.1.2 European Region

```{r}
water_long_selected_reshape_10 <- water_long_selected_reshape %>%
  select('Country', 'Region',`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  filter(Region == 'European Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_10 <- water_long_selected_reshape_10 %>%
  drop_na(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
          `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  mutate(
    `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources` = 
      round(as.numeric(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`), 0),
    
    `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)` = 
      round(as.numeric(`Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`), 0)
  )

```

```{r}
p2 <- ggplot(data = water_long_selected_reshape_10, 
             aes(x = `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
                 y = `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(0, 140), ylim = c(0, 140)) +
  labs(x = "Level of water stress (%)", 
       y = "Water Productivity (US$ GDP per m³)",
       color = "Country Name")
p2
```

```{r}
girafe(ggobj = p2)
```

### 3.1.3 Western Pacific Region

```{r}
water_long_selected_reshape_11 <- water_long_selected_reshape %>%
  select('Country', 'Region',`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  filter(Region == 'Western Pacific Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_11 <- water_long_selected_reshape_11 %>%
  drop_na(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
          `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  mutate(
    `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources` = 
      round(as.numeric(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`), 0),
    
    `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)` = 
      round(as.numeric(`Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`), 0)
  )

```

```{r}
p3 <- ggplot(data = water_long_selected_reshape_11, 
             aes(x = `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
                 y = `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(0, 140), ylim = c(0, 140)) +
  labs(x = "Level of water stress (%)", 
       y = "Water Productivity (US$ GDP per m³)",
       color = "Country Name")

p3
```

```{r}
girafe(ggobj = p3)
```

### 3.1.4 Eastern Mediterranean Region

```{r}
water_long_selected_reshape_12 <- water_long_selected_reshape %>%
  select('Country', 'Region',`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  filter(Region == 'Eastern Mediterranean Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_12 <- water_long_selected_reshape_12 %>%
  drop_na(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
          `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  mutate(
    `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources` = 
      round(as.numeric(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`), 0),
    
    `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)` = 
      round(as.numeric(`Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`), 0)
  )

```

```{r}
p4 <- ggplot(data = water_long_selected_reshape_12, 
             aes(x = `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
                 y = `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(0, 140), ylim = c(0, 140)) +
  labs(x = "Level of water stress (%)", 
       y = "Water Productivity (US$ GDP per m³)",
       color = "Country Name")

p4
```

```{r}
girafe(ggobj = p4)
```

### 3.1.5 South-East Asia Region

```{r}
water_long_selected_reshape_13 <- water_long_selected_reshape %>%
  select('Country', 'Region',`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  filter(Region == 'South-East Asia Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_13 <- water_long_selected_reshape_13 %>%
  drop_na(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
          `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  mutate(
    `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources` = 
      round(as.numeric(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`), 0),
    
    `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)` = 
      round(as.numeric(`Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`), 0)
  )

```

```{r}
p5 <- ggplot(data = water_long_selected_reshape_13, 
             aes(x = `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
                 y = `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(0, 140), ylim = c(0, 140)) +
  labs(x = "Level of water stress (%)", 
       y = "Water Productivity (US$ GDP per m³)",
       color = "Country Name")

p5
```

```{r}
girafe(ggobj = p5)
```

### 3.1.6 Region of the Americas

```{r}
water_long_selected_reshape_14 <- water_long_selected_reshape %>%
  select('Country', 'Region',`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  filter(Region == 'Region of the Americas') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_14 <- water_long_selected_reshape_14 %>%
  drop_na(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
          `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`) %>%
  mutate(
    `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources` = 
      round(as.numeric(`Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`), 0),
    
    `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)` = 
      round(as.numeric(`Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`), 0)
  )

```

```{r}
p6 <- ggplot(data = water_long_selected_reshape_14, 
             aes(x = `Level of water stress: freshwater withdrawal as a proportion of available freshwater resources`, 
                 y = `Water productivity, total (constant 2015 US$ GDP per cubic meter of total freshwater withdrawal)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(0, 140), ylim = c(0, 140)) +
  labs(x = "Level of water stress (%)", 
       y = "Water Productivity (US$ GDP per m³)",
       color = "Country Name")

p6
```

```{r}
girafe(ggobj = p6)
```

## 3.2 Simple scatterplot between basic drinking water services (%) and safely managed drinking water services (%) by region

### 3.2.1 African Region

```{r}
water_long_selected_reshape_15 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  filter(Region == 'African Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_15 <- water_long_selected_reshape_15 %>%
  drop_na(`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  mutate(
    `People using at least basic drinking water services (% of population)` = 
      round(as.numeric(`People using at least basic drinking water services (% of population)`), 0),
    
    `People using safely managed drinking water services (% of population)` = 
      round(as.numeric(`People using safely managed drinking water services (% of population)`), 0)
  )

```

```{r}
p7 <- ggplot(data = water_long_selected_reshape_15, 
             aes(x = `People using at least basic drinking water services (% of population)`, 
                 y = `People using safely managed drinking water services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(30, 100), ylim = c(5, 100)) +
  labs(x = "Basic drinking water services (%)", 
       y = "safely managed drinking water services",
       color = "Country Name")
p7
```

```{r}
girafe(ggobj = p7)
```

### 3.2.2 European Region

```{r}
water_long_selected_reshape_16 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  filter(Region == 'European Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_16 <- water_long_selected_reshape_16 %>%
  drop_na(`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  mutate(
    `People using at least basic drinking water services (% of population)` = 
      round(as.numeric(`People using at least basic drinking water services (% of population)`), 0),
    
    `People using safely managed drinking water services (% of population)` = 
      round(as.numeric(`People using safely managed drinking water services (% of population)`), 0)
  )

```

```{r}
p8 <- ggplot(data = water_long_selected_reshape_16, 
             aes(x = `People using at least basic drinking water services (% of population)`, 
                 y = `People using safely managed drinking water services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(30, 100), ylim = c(5, 100)) +
  labs(x = "Basic drinking water services (%)", 
       y = "safely managed drinking water services",
       color = "Country Name")
p8
```

```{r}
girafe(ggobj = p8)
```

### 3.2.3 Western Pacific Region

```{r}
water_long_selected_reshape_17 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  filter(Region == 'European Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_17 <- water_long_selected_reshape_17 %>%
  drop_na(`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  mutate(
    `People using at least basic drinking water services (% of population)` = 
      round(as.numeric(`People using at least basic drinking water services (% of population)`), 0),
    
    `People using safely managed drinking water services (% of population)` = 
      round(as.numeric(`People using safely managed drinking water services (% of population)`), 0)
  )

```

```{r}
p9 <- ggplot(data = water_long_selected_reshape_17, 
             aes(x = `People using at least basic drinking water services (% of population)`, 
                 y = `People using safely managed drinking water services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(30, 100), ylim = c(5, 100)) +
  labs(x = "Basic drinking water services (%)", 
       y = "safely managed drinking water services",
       color = "Country Name")
p9
```

```{r}
girafe(ggobj = p9)
```

### 3.2.4 Eastern Mediterranean Region

```{r}
water_long_selected_reshape_18 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  filter(Region == 'Eastern Mediterranean Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_18 <- water_long_selected_reshape_18 %>%
  drop_na(`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  mutate(
    `People using at least basic drinking water services (% of population)` = 
      round(as.numeric(`People using at least basic drinking water services (% of population)`), 0),
    
    `People using safely managed drinking water services (% of population)` = 
      round(as.numeric(`People using safely managed drinking water services (% of population)`), 0)
  )

```

```{r}
p10 <- ggplot(data = water_long_selected_reshape_18, 
             aes(x = `People using at least basic drinking water services (% of population)`, 
                 y = `People using safely managed drinking water services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(30, 100), ylim = c(5, 100)) +
  labs(x = "Basic drinking water services (%)", 
       y = "safely managed drinking water services",
       color = "Country Name")
p10
```

```{r}
girafe(ggobj = p10)
```

### 3.2.5 South-East Asia Region

```{r}
water_long_selected_reshape_19 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  filter(Region == 'South-East Asia Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_19 <- water_long_selected_reshape_19 %>%
  drop_na(`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  mutate(
    `People using at least basic drinking water services (% of population)` = 
      round(as.numeric(`People using at least basic drinking water services (% of population)`), 0),
    
    `People using safely managed drinking water services (% of population)` = 
      round(as.numeric(`People using safely managed drinking water services (% of population)`), 0)
  )

```

```{r}
p11 <- ggplot(data = water_long_selected_reshape_19, 
             aes(x = `People using at least basic drinking water services (% of population)`, 
                 y = `People using safely managed drinking water services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(30, 100), ylim = c(5, 100)) +
  labs(x = "Basic drinking water services (%)", 
       y = "safely managed drinking water services",
       color = "Country Name")
p11
```

```{r}
girafe(ggobj = p11)
```

### 3.2.6 Region of the Americas

```{r}
water_long_selected_reshape_20 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  filter(Region == 'Region of the Americas') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_20 <- water_long_selected_reshape_20 %>%
  drop_na(`People using at least basic drinking water services (% of population)`, `People using safely managed drinking water services (% of population)`) %>%
  mutate(
    `People using at least basic drinking water services (% of population)` = 
      round(as.numeric(`People using at least basic drinking water services (% of population)`), 0),
    
    `People using safely managed drinking water services (% of population)` = 
      round(as.numeric(`People using safely managed drinking water services (% of population)`), 0)
  )

```

```{r}
p12 <- ggplot(data = water_long_selected_reshape_20, 
             aes(x = `People using at least basic drinking water services (% of population)`, 
                 y = `People using safely managed drinking water services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(30, 100), ylim = c(5, 100)) +
  labs(x = "Basic drinking water services (%)", 
       y = "safely managed drinking water services",
       color = "Country Name")
p12
```

```{r}
girafe(ggobj = p12)
```

## 3.3 Simple scatterplot between basic sanitation services (% of population) and safely managed sanitation services (% of population)

### 3.3.1 African Region

```{r}
water_long_selected_reshape_21 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  filter(Region == 'African Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_21 <- water_long_selected_reshape_21 %>%
  drop_na(`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  mutate(
    `People using at least basic sanitation services (% of population)` = 
      round(as.numeric(`People using at least basic sanitation services (% of population)`), 0),
    
    `People using safely managed sanitation services (% of population)` = 
      round(as.numeric(`People using safely managed sanitation services (% of population)`), 0)
  )

```

```{r}
p13 <- ggplot(data = water_long_selected_reshape_21, 
             aes(x = `People using at least basic sanitation services (% of population)`, 
                 y = `People using safely managed sanitation services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(5, 100), ylim = c(0, 100)) +
  labs(x = "Basic sanitation services (%)", 
       y = "safely managed sanitation services (%)",
       color = "Country Name")
p13
```

```{r}
girafe(ggobj = p13)
```

### 3.3.2 European Region

```{r}
water_long_selected_reshape_22 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  filter(Region == 'European Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_22 <- water_long_selected_reshape_22 %>%
  drop_na(`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  mutate(
    `People using at least basic sanitation services (% of population)` = 
      round(as.numeric(`People using at least basic sanitation services (% of population)`), 0),
    
    `People using safely managed sanitation services (% of population)` = 
      round(as.numeric(`People using safely managed sanitation services (% of population)`), 0)
  )

```

```{r}
p14 <- ggplot(data = water_long_selected_reshape_22, 
             aes(x = `People using at least basic sanitation services (% of population)`, 
                 y = `People using safely managed sanitation services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(5, 100), ylim = c(0, 100)) +
  labs(x = "Basic sanitation services (%)", 
       y = "safely managed sanitation services (%)",
       color = "Country Name")
p14
```

```{r}
girafe(ggobj = p14)
```

### 3.3.3 Western Pacific Region

```{r}
water_long_selected_reshape_23 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  filter(Region == 'Western Pacific Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_23 <- water_long_selected_reshape_23 %>%
  drop_na(`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  mutate(
    `People using at least basic sanitation services (% of population)` = 
      round(as.numeric(`People using at least basic sanitation services (% of population)`), 0),
    
    `People using safely managed sanitation services (% of population)` = 
      round(as.numeric(`People using safely managed sanitation services (% of population)`), 0)
  )

```

```{r}
p15 <- ggplot(data = water_long_selected_reshape_23, 
             aes(x = `People using at least basic sanitation services (% of population)`, 
                 y = `People using safely managed sanitation services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(5, 100), ylim = c(0, 100)) +
  labs(x = "Basic sanitation services (%)", 
       y = "safely managed sanitation services (%)",
       color = "Country Name")
p15
```

```{r}
girafe(ggobj = p15)
```

### 3.3.4 Eastern Mediterranean Region

```{r}
water_long_selected_reshape_24 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  filter(Region == 'Eastern Mediterranean Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_24 <- water_long_selected_reshape_24 %>%
  drop_na(`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  mutate(
    `People using at least basic sanitation services (% of population)` = 
      round(as.numeric(`People using at least basic sanitation services (% of population)`), 0),
    
    `People using safely managed sanitation services (% of population)` = 
      round(as.numeric(`People using safely managed sanitation services (% of population)`), 0)
  )

```

```{r}
p16 <- ggplot(data = water_long_selected_reshape_24, 
             aes(x = `People using at least basic sanitation services (% of population)`, 
                 y = `People using safely managed sanitation services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(5, 100), ylim = c(0, 100)) +
  labs(x = "Basic sanitation services (%)", 
       y = "safely managed sanitation services (%)",
       color = "Country Name")
p16
```

```{r}
girafe(ggobj = p16)
```

### 3.3.5 South-East Asia Region

```{r}
water_long_selected_reshape_25 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  filter(Region == 'South-East Asia Region') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_25 <- water_long_selected_reshape_25 %>%
  drop_na(`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  mutate(
    `People using at least basic sanitation services (% of population)` = 
      round(as.numeric(`People using at least basic sanitation services (% of population)`), 0),
    
    `People using safely managed sanitation services (% of population)` = 
      round(as.numeric(`People using safely managed sanitation services (% of population)`), 0)
  )

```

```{r}
p17 <- ggplot(data = water_long_selected_reshape_24, 
             aes(x = `People using at least basic sanitation services (% of population)`, 
                 y = `People using safely managed sanitation services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(5, 100), ylim = c(0, 100)) +
  labs(x = "Basic sanitation services (%)", 
       y = "safely managed sanitation services (%)",
       color = "Country Name")
p17
```

```{r}
girafe(ggobj = p17)
```

### 3.3.6 Region of the Americas

```{r}
water_long_selected_reshape_26 <- water_long_selected_reshape %>%
  select('Country', 'Region',`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  filter(Region == 'Region of the Americas') %>%
  mutate(Country = as.factor(Country))
```

```{r}
water_long_selected_reshape_26 <- water_long_selected_reshape_26 %>%
  drop_na(`People using at least basic sanitation services (% of population)`, `People using safely managed sanitation services (% of population)`) %>%
  mutate(
    `People using at least basic sanitation services (% of population)` = 
      round(as.numeric(`People using at least basic sanitation services (% of population)`), 0),
    
    `People using safely managed sanitation services (% of population)` = 
      round(as.numeric(`People using safely managed sanitation services (% of population)`), 0)
  )

```

```{r}
p18 <- ggplot(data = water_long_selected_reshape_26, 
             aes(x = `People using at least basic sanitation services (% of population)`, 
                 y = `People using safely managed sanitation services (% of population)`, 
                 color = Country)) +  
  geom_point_interactive(aes(
    tooltip = Country))+
  coord_cartesian(xlim = c(5, 100), ylim = c(0, 100)) +
  labs(x = "Basic sanitation services (%)", 
       y = "safely managed sanitation services (%)",
       color = "Country Name")
p18
```

```{r}
girafe(ggobj = p18)
```
